Demographic Influences on Municipal Budget Allocation and Infrastructure Development in Toronto

### 2024-12-04 paper/paper.qmd
Updated Files
- Added Datasheet
- Added LLM usage
- Final checks in paper
- Updated reference.bib

---
title: "Demographic Influences on Municipal Budget Allocation and Infrastructure Development in Toronto"
subtitle: "Population Density Drive Larger Budget Allocations meanwhile Average Household Income Drives Construction Projects"
author: 
  - Aviral Bhardwaj
thanks: "Code and data are available at: https://github.com/Aviral-03/Toronto-WardWide-Budget-Analysis"
date: today
date-format: long
abstract: "This study investigates how population density and average household income affect budget allocations across Toronto’s 25 wards, and how these factors, along with budget allocation, influence construction projects. Using data from the city's \"Capital Budget Plan\", \"Ward Profiles\" and \"Active Building Permits\", we analyze spending patterns in key areas like related to infrastructre. Our analysis, guided by a causal model, reveals a negative association between population density and construction activity, with some high-density wards receiving significantly higher budget allocations. Average household income and total budget allocations were positively correlated with the number of building permits. These insights highlight potential disparities in resource distribution, emphasizing the need for more equitable urban budget planning to address diverse community needs."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
prefer-html: true
---

@@

The paper is organized as follows: [@sec-data] outlines the data sources and methodology used in the analysis. [@sec-model] explains the Bayesian model applied to estimate the effects of demographic and economic factors, along with budget allocations, on construction projects. [@sec-results] presents key findings, including trends in budget allocations, the relationship between demographic factors and budget distribution, and the impact on construction projects. [@sec-discussion] explores the implications for urban governance and budget planning in Toronto. Finally, [@sec-limitations] addresses the study’s limitations and [@sec-future-work] offers recommendations for future research. Additional details, including tables, figures, and methodology, are provided in the [@sec-appendix].

@@

Population is a critical demographic indicator, representing the number of residents in each ward, while average household income reflects economic well-being, influencing access to resources and quality of life [@Schaeffer]. Therefore, these two variables of interest were chosen for the analysis. For budget allocations we choose these specific categories of interest, specifically Growth-Related expenditures, State of Good Repair, and Service Improvement and Enhancement.

@@

This data-set was included in this analysis to provide insights into the population and average household income for each ward, providing insights into the city's socioeconomic landscape. 25-Ward model was used instead of the 44-Ward model as it was the most recent data available at the time of analysis and matched the Capital Budget data.

@@

These categories were selected because they represent critical areas of investment that directly impact the quality of life and well-being of Toronto residents. *State of Good Repair* focuses on maintaining and preserving existing infrastructure, *Growth Related* addresses the expansion and development of new infrastructure, and *Service Improvement and Enhancement* aims to enhance public services and amenities. By analyzing budget allocations in these categories, we can gain valuable insights into the city's spending priorities and identify opportunities for more effective and equitable resource distribution. While also identifying potential disparities in funding across various wards of the city.

@@

The *Building Permits - Active Permits* dataset [@toronto_building_permits], published by the City Planning Division and last updated on November 28, 2024, serves as a comprehensive record of active building permits in Toronto. A building permit is a municipally issued document, mandated by the Building Code Act and enforced by the City of Toronto, regulating the construction or demolition of physical structures [@toronto_building_permits].

@@

We selected this data-set to evaluate how budget allocations correlate with the number of building permits issued in each ward. The number of permits serves as a proxy for construction activity and development, highlighting the demand for infrastructure investment and capital expenditures. By examining the relationship between building permits and budget allocations, we can gain valuable insights into how construction activity influences resource distribution across the city's wards.

@@

[@fig-building-permits-data] illustrates the total number of building permits by ward, providing valuable insights into construction activity and development patterns across Toronto’s wards.

@@

average_budget |>
  head() |>
  kable(col.names = c("Ward ID", "Ward Name", "Growth Related", 
                      "State of Good Repair", "Service Improvement and Enhancement"))
```
### Relationship between Average Household Income & Population with Budget Allocation for Service Improvement and Enhancement Projects 
```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-service-improvement-budget
#| fig.cap: "Service Improvement and Enhancement Budget by Ward"
#| fig-pos: "h"
#| fig.width: 10
#| fig.height: 5

@@

### Relationship between Average Household Income & Population with Budget Allocation for Growth Related Projects and State of Good Repair Projects

[@fig-growth-related-budget] (a) illustrates the relationship between average household income, population, and the total 10-year budget allocation for Growth-Related projects. The linear regression line for average household income remains relatively flat, hovering around \$100,000. This suggests that budget allocations for Growth-Related projects are not significantly influenced by income levels, indicating a more uniform distribution of funds across wards regardless of household income. A similar lack of correlation is observed in [@fig-growth-related-budget] (b), where budget allocations for State of Good Repair projects also show minimal dependence on average household income.

In contrast, the linear regression line for population shows a strong positive correlation with budget allocation for Growth-Related projects. Wards with higher population densities—such as Ward 10 (Spadina-Fort York), Ward 13 (Toronto Centre), Ward 5 (York South-Weston), and Ward 2 (Etobicoke North)—receive a larger share of the Growth-Related budget. This indicates that population density plays a critical role in determining funding levels for these projects, with more populous wards benefiting from increased allocations. A similar trend is observed in the budget distribution for State of Good Repair projects, where wards with higher population densities receive a proportionally larger share of the funds.
```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-growth-related-budget
#| fig.cap: "Growth Related Budget and State of Good Repair Budget by Ward"
#| fig-subcap: ["Growth Related Budget", "State of Good Repair Budget"]
#| fig-pos: "h"
#| fig.width: 10
#| fig.height: 5
#| layout-ncol: 1

@@

p2 <- ggplot(growth_budget, aes(x = average_household_income, y = `Growth Related`, label= ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(x = "Average Household Income", y = "") +
  theme_minimal()

# Combine both plots side by side
grid.arrange(p1, p2, ncol = 2)

good_repair_budget <- combined_data |>
  select(ward_id, ward, population, average_household_income, `State of Good Repair`)

p1 <- ggplot(good_repair_budget, aes(x = population, y = `State of Good Repair`, label= ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(x = "Population", y = "Average Total 10-Year Budget (in 000s) 2021-24") +
  theme_minimal()

p2 <- ggplot(good_repair_budget, aes(x = average_household_income, y = `State of Good Repair`, label= ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(x = "Average Household Income", y = "") +
  theme_minimal()

# Combine both plots side by side
grid.arrange(p1, p2, ncol = 2)

@@

To better understand the underlying data, [@eda-model-justification] provides a summary of the mean, median, and standard deviation for each variable used in the model. The predictor variables have large values and different scales, which can cause numerical instability and convergence issues in the model. To address this, we will standardize the predictor variables by subtracting the mean and dividing by the standard deviation. This will scale the variables consistently and improve both the model's convergence and interpretability. Additionally, we did not use a Bayesian Poisson regression model or negative binomial because the number of building permits is a continuous variable, not count data. Therefore, we chose to use a Bayesian multiple linear regression model. This model allows us to estimate the effect of population density, average household income, and budget allocations on the number of building permits issued in each ward giving us a well-rounded understanding of the relationships between these variables.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: eda-model-justification
#| tbl-cap: "Mean, Median, and Standard Deviation of Each Variable"
#| tbl-pos: "h"

@@

@@

Population distribution reveals a concentration in the western part of the city and downtown areas. Notably, wards like Etobicoke Centre (Ward 2) and Etobicoke-Lakeshore (Ward 3), situated farther from downtown, have lower costs of living and correspondingly lower average household incomes, such as Etobicoke North (Ward 2) with \$95,200.

@@

The results of the Bayesian regression model, summarized in [@tbl-model-results], provide important insights into the relationship between total budget allocations, average household income, and the number of building permits issued across wards.

The model explains 59.1% of the variance in building permits (R² = 0.591), indicating a moderate level of explanatory power. The coefficients suggest that both total budget (0.530) and average household income (0.350) have a positive and significant impact on the number of building permits issued, implying that areas with higher budget allocations and income levels are more likely to see increased construction activity.

The inclusion of ward-specific random effects (σ² = 0.141) highlights the importance of accounting for variation between wards, suggesting that there are unobserved factors specific to each ward influencing permit issuance. Although the adjusted R² (0.369) indicates that the model's explanatory power is somewhat limited after accounting for the complexity of the predictors, the model still provides valuable insights into the role of economic factors in shaping construction activity. With an RMSE of 0.58, the model's predictions are reasonably accurate, although there is still room for improvement.

@@

As highlighted in [@sec-model-results] Bayesian regression model reveals that average household income has a significant positive effect on the issuance of building permits. This suggests that construction activity tends to concentrate in higher-income areas, where demand for development is stronger. This finding underscores the role of income levels in driving construction and infrastructure development, emphasizing the need for targeted investments in lower-income areas to stimulate economic growth and enhance community well-being.

For potential property investors in Toronto, these insights offer a strategic advantage. Investing in areas with higher average household incomes may be more profitable due to the increased likelihood of ongoing construction activity, which can drive property value appreciation and strengthen the local economy. While high-population-density areas are typically attractive for investment, our findings suggest that this may not always be the case.

@@

These findings underscore the importance of equity-based budgeting, where low-density and lower-income areas should receive targeted investments to meet critical infrastructure needs. This approach aligns with urban development strategies that prioritize inclusivity. Notable examples include transit expansions such as the Ontario Line and Crosstown extensions [@citeTransitExpansion].The Ontario Line is expected to connect high-density areas like Ward 10 (Spadina-Fort York) and Ward 13 (Toronto Centre) to the broader transit network, which could stimulate further development and infrastructure investment in these wards, fostering a more balanced urban landscape and distributing population growth more evenly across the city.

@@

# Weaknesses and Limitations {#sec-limitations}

There are several limitations to the analysis that need to be addressed. As discussed in [@sec-data], the datasets used in this study are publicly available from the City of Toronto. While they provide valuable insights into the city's demographics, budget allocations, and building permits, they may not fully capture the complexity of Toronto’s infrastructure development and resource distribution. For example, the building permits dataset includes only active permits, which may not reflect the entire scope of construction activity. Additionally, the budget allocations dataset may exclude certain infrastructure projects and expenditures, leading to potential gaps in the analysis.

Furthermore, as detailed in [@sec-appendix], there are 45 unique statuses assigned to each submitted permit. This analysis focused on only one subset of these statuses, which may not fully capture the complexity of the building permit process and the various stages of construction. Future research could include additional permit statuses for a more comprehensive understanding of construction activity. Another limitation is the lack of detailed data on specific projects within each budget category. This information could offer further insights into why certain wards receive more or less funding.

@@

# Future Work {#sec-future-work}

This analysis provides a foundation for future research on infrastructure development, resource distribution, and economic well-being in Toronto. Future studies could explore additional factors that influence budget allocations and building permit issuance, such as infrastructure needs, community priorities, and political considerations. By incorporating these variables into the analysis, researchers can gain a more comprehensive understanding of how municipal budgets are allocated and how construction activity is distributed across the city's wards.

@@

@@

>>>>

---
title: "Demographic Influences on Municipal Budget Allocation and Infrastructure Development in Toronto"
subtitle: "Population Density Drive Larger Budget Allocations meanwhile Average Household Income Drives Construction Projects"
author: 
  - Aviral Bhardwaj
thanks: "Code and data are available at: [https://github.com/Aviral-03/InfrastructureCausalModel-TO](https://github.com/Aviral-03/InfrastructureCausalModel-TO)." 
date: today
date-format: long
abstract: "This study investigates how population density and average household income affect budget allocations across Toronto’s 25 wards, and how these factors, along with budget allocation, influence construction projects. Using data from the city's \"Capital Budget Plan\", \"Ward Profiles\" and \"Active Building Permits\", we analyze spending patterns in key areas like related to infrastructre. Our analysis, guided by a causal model, presents a negative association between population density and construction activity, with some high-density wards receiving significantly higher budget allocations. Average household income and total budget allocations were positively correlated with the number of building permits. These findings highlight potential disparities in resource distribution, emphasizing the need for more equitable urban budget planning to address diverse community needs."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
prefer-html: true
---

@@

The paper is organized as follows: [@sec-data] outlines the data sources and methodology used in the analysis. [@sec-model] explains the Bayesian model applied to estimate the effects of demographic and economic factors, along with budget allocations, on construction projects. [@sec-results] presents key findings, including trends in budget allocations, the relationship between demographic factors and budget distribution, and the impact on construction projects. [@sec-discussion] explores the implications for urban governance and budget planning in Toronto. Finally, [@sec-limitations] addresses the study’s limitations and offers recommendations for future research. Additional details, including tables, figures, and methodology, are provided in the [@sec-appendix].

@@

Population is a important demographic indicator, representing the number of residents in each ward, while average household income reflects economic well-being, influencing access to resources and quality of life [@Schaeffer]. Therefore, these two variables of interest were chosen for the analysis. For budget allocations we choose these specific categories of interest, specifically Growth-Related expenditures, State of Good Repair, and Service Improvement and Enhancement.

@@

This data-set was included in this analysis to provide understanding into the population and average household income for each ward, providing understanding into the city's socioeconomic landscape. 25-Ward model was used instead of the 44-Ward model as it was the most recent data available at the time of analysis and matched the Capital Budget data.

@@

These categories were selected because they represent significant areas of investment that directly impact the quality of life and well-being of Toronto residents. *State of Good Repair* focuses on maintaining and preserving existing infrastructure, *Growth Related* addresses the expansion and development of new infrastructure, and *Service Improvement and Enhancement* aims to enhance public services and amenities. By analyzing budget allocations in these categories, we can gain understand city's spending priorities and identify opportunities for more effective and equitable resource distribution. While also identifying potential disparities in funding across various wards of the city.

@@

The *Building Permits - Active Permits* dataset [@toronto_building_permits], published by the City Planning Division and last updated on November 28, 2024, serves as an important record of active building permits in Toronto. A building permit is a municipally issued document, mandated by the Building Code Act and enforced by the City of Toronto, regulating the construction or demolition of physical structures [@toronto_building_permits].

@@

We selected this data-set to evaluate how budget allocations correlate with the number of building permits issued in each ward. The number of permits serves as a proxy for construction activity and development, highlighting the demand for infrastructure investment and capital expenditures. By examining the relationship between building permits and budget allocations, we can understand how construction activity influences resource distribution across the city's wards.

@@

[@fig-building-permits-data] illustrates the total number of building permits by ward, providing understanding into construction activity and development patterns across Toronto’s wards.

@@

average_budget |>
  head() |>
  kable(col.names = c("Ward ID", "Ward Name", "Growth Related", 
                      "State of Good Repair", "Service Improvement and Enhancement"))
```

### Relationship between Predictors and Budget Allocations (Service Improvement and Enhancement)
```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-service-improvement-budget
#| fig.cap: "Service Improvement and Enhancement Budget by Ward"
#| fig-pos: "h"
#| fig.width: 10
#| fig.height: 5

@@

### Relationship between Predictors and Budget Allocations (Growth-Related and State of Good Repair)

[@fig-growth-related-budget]a) illustrates the relationship between average household income, population, and the total 10-year budget allocation for Growth-Related projects. The linear regression line for average household income remains relatively flat, hovering around \$100,000. This suggests that budget allocations for Growth-Related projects are not significantly influenced by income levels, indicating a more uniform distribution of funds across wards regardless of household income. A similar lack of correlation is observed in [@fig-growth-related-budget]b), where budget allocations for State of Good Repair projects also show minimal dependence on average household income.

In contrast, the linear regression line for population shows a strong positive correlation with budget allocation for Growth-Related projects. Wards with higher population densities—such as Ward 10 (Spadina-Fort York), Ward 13 (Toronto Centre), Ward 5 (York South-Weston), and Ward 2 (Etobicoke North)—receive a larger share of the Growth-Related budget. This indicates that population density plays an important role in determining funding levels for these projects, with more populous wards benefiting from increased allocations. A similar trend is observed in the budget distribution for State of Good Repair projects, where wards with higher population densities receive a proportionally larger share of the funds.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-growth-related-budget
#| fig.cap: "Growth and State of Good Repair Budget by Ward"
#| fig-subcap: ["Growth Related", "State of Good Repair"]
#| fig-pos: "h"
#| fig.width: 10
#| fig.height: 3

@@

p2 <- ggplot(growth_budget, aes(x = average_household_income, y = `Growth Related`, label= ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(x = "Average Household Income", y = "") +
  theme_minimal()
good_repair_budget <- combined_data |>
  select(ward_id, ward, population, average_household_income, `State of Good Repair`)

p3 <- ggplot(good_repair_budget, aes(x = population, y = `State of Good Repair`, label= ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(x = "Population", y = "Average Total 10-Year Budget (in 000s) 2021-24") +
  theme_minimal()

p4 <- ggplot(good_repair_budget, aes(x = average_household_income, y = `State of Good Repair`, label= ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(x = "Average Household Income", y = "") +
  theme_minimal()

# Combine both plots side by side
grid.arrange(p1, p2, ncol = 2)
grid.arrange(p3, p4, ncol = 2)

# | echo: false
# | message: false
# | warning: false
# | label: fig-good-repair-budget
# | fig.cap: "State of Good Repair Budget by Ward"
# | fig-pos: "h"
# | fig.width: 10
# | fig.height: 5
```

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-model-justification
#| tbl-cap: "Mean, Median, and Standard Deviation of Each Variable"
#| tbl-pos: "h"

@@

To better understand the underlying data, [@tbl-model-justification] provides a summary of the mean, median, and standard deviation for each variable used in the model. The predictor variables have large values and different scales, which can cause numerical instability and convergence issues in the model. To address this, we will standardize the predictor variables by subtracting the mean and dividing by the standard deviation. This will scale the variables consistently and improve both the model's convergence and interpretability. Additionally, we did not use a Bayesian Poisson regression model or negative binomial because the number of building permits is a continuous variable, not count data. Therefore, we chose to use a Bayesian multiple linear regression model. This model allows us to estimate the effect of population density, average household income, and budget allocations on the number of building permits issued in each ward giving us a well-rounded understanding of the relationships between these variables.

@@

Population distribution presents a concentration in the western part of the city and downtown areas. Notably, wards like Etobicoke Centre (Ward 2) and Etobicoke-Lakeshore (Ward 3), situated farther from downtown, have lower costs of living and correspondingly lower average household incomes, such as Etobicoke North (Ward 2) with \$95,200.

@@

The results of the Bayesian regression model, summarized in [@tbl-model-results], provide important understanding into the relationship between total budget allocations, average household income, and the number of building permits issued across wards.

The model explains 59.1% of the variance in building permits, indicating a moderate level of explanatory power. The coefficients suggest that both total budget (0.530) and average household income (0.350) have a positive and significant impact on the number of building permits issued, implying that areas with higher budget allocations and income levels are more likely to see increased construction activity.

The inclusion of ward-specific random effects highlights the importance of accounting for variation between wards, suggesting that there are unobserved factors specific to each ward influencing permit issuance. Although the adjusted $R^2$ (0.369) indicates that the model's explanatory power is somewhat limited after accounting for the complexity of the predictors, the model still provides understanding into the role of economic factors in shaping construction activity. With an RMSE of 0.58, the model's predictions are reasonably accurate, although there is still room for improvement.

@@

As highlighted in [@sec-model-results] Bayesian regression model shows that average household income has a significant positive effect on the issuance of building permits. This suggests that construction activity tends to concentrate in higher-income areas, where demand for development is stronger. This finding underscores the role of income levels in driving construction and infrastructure development, emphasizing the need for targeted investments in lower-income areas to stimulate economic growth and enhance community well-being.

For potential property investors in Toronto, these understanding offer a strategic advantage. Investing in areas with higher average household incomes may be more profitable due to the increased likelihood of ongoing construction activity, which can drive property value appreciation and strengthen the local economy. While high-population-density areas are typically attractive for investment, our findings suggest that this may not always be the case.

@@

These findings underscore the importance of equity-based budgeting, where low-density and lower-income areas should receive targeted investments to meet infrastructure needs. This approach aligns with urban development strategies that prioritize inclusivity. Notable examples include transit expansions such as the Ontario Line and Crosstown extensions [@citeTransitExpansion].The Ontario Line is expected to connect high-density areas like Ward 10 (Spadina-Fort York) and Ward 13 (Toronto Centre) to the broader transit network, which could stimulate further development and infrastructure investment in these wards, fostering a more balanced urban landscape and distributing population growth more evenly across the city.

@@

# Weaknesses and Future Work {#sec-limitations}

There are several limitations to the analysis that need to be addressed. As discussed in [@sec-data], the datasets used in this study are publicly available from the City of Toronto. While they represents city's demographics, budget allocations, and building permits, they may not fully capture the complexity of Toronto’s infrastructure development and resource distribution. For example, the building permits dataset includes only active permits, which may not reflect the entire scope of construction activity. Additionally, the budget allocations dataset may exclude certain infrastructure projects and expenditures, leading to potential gaps in the analysis.

Furthermore, as detailed in [@sec-appendix], there are 45 unique statuses assigned to each submitted permit. This analysis focused on only one subset of these statuses, which may not fully capture the complexity of the building permit process and the various stages of construction. Future research could include additional permit statuses for a better understanding of construction activity. Another limitation is the lack of detailed data on specific projects within each budget category. This information could offer further understanding into why certain wards receive more or less funding.

@@

## Future Work

This analysis provides a foundation for future research on infrastructure development, resource distribution, and economic well-being in Toronto. Future studies could explore additional factors that influence budget allocations and building permit issuance, such as infrastructure needs, community priorities, and political considerations. By incorporating these variables into the analysis, researchers can gain a more in-depth understanding of how municipal budgets are allocated and how construction activity is distributed across the city's wards.

@@

\appendix

@@

## Survey, Sample, and Data Collection

The relationship between income levels, population and construction outcomes is complex and influenced by diverse socioeconomic and regional factors. Current observational data, such as NHS survey to get income levels, and population of ward profiles but is inherently limited by its sample [@nhs_canada]. To better answer our research question, a more targeted survey approach is essential. This exploration outlines an idealized survey methodology designed to reduce bias, ensure representativeness, and deepen understanding.

### Idealized Survey Design for Construction Analysis 
A survey specifically designed to explore the link between income and construction outcomes would involve the following key components:

a. Defining the Survey Objectives
The survey would aim to investigate how income disparities influence construction investment, access to resources, regulatory hurdles, and housing quality across different regions.

b. Sampling Methodology  
To ensure a representative and unbiased dataset, the survey would employ:

- Stratified Sampling: Income levels and ward classifications (e.g., urban, suburban, rural) would be stratified to ensure proportional representation.  
- Random Sampling Within Strata: Participants would be randomly selected within each stratum to enhance the validity and generalizability of the findings.

c. Sample Size and Distribution
The sample size should be sufficiently large to allow for statistically significant comparisons across different income levels and geographic regions. A minimum target of several thousand participants distributed evenly across strata would be ideal.

### Reducing Bias in Sampling and Data Collection  
Bias reduction is important in survey design to ensure reliable and actionable understanding. This involves addressing various types of bias:

a. Sampling Bias Mitigation
Stratified random sampling ensures that underrepresented groups, particularly those in lower income brackets or remote regions, are adequately included.

b. Non-Response Bias Mitigation 
Offering multiple survey delivery methods—such as online platforms, in-person interviews, and mailed surveys—enhances accessibility and participation. Additionally, providing incentives like small monetary rewards or gift cards can boost response rates among diverse groups.

c. Design Bias Mitigation 
To avoid influencing responses, survey questions must be neutrally worded and pilot-tested to ensure clarity. Cognitive interviews with a subset of participants can help refine the wording and structure of questions.
